Unsupervised embedding of trajectories captures the latent structure of mobility

### 2021-02-18 paper/current/main.tex
edit math section

Why does the embedding distance predict actual mobility flows better than the geographic distance?
Here, we show that the objective function of {\it word2vec} is equivalent to the gravity model.

The {\it word2vec} model takes a location trajectory, denoted by ($a_{1}, a_{2}, \ldots, a_{T}$), as an input.
A location $a_t=i$ is considered to have a context location $a_{t'}=j$ that appears in its succeeding or proceeding $w$ locations in the trajectory, \ie  $j \in [a_{t-w}, \ldots, a_{t-1}, a_{t+1},\ldots, a_{t+w}]$.
	{\it Word2vec} learns an embedding by estimating the probability that location $i$ has context $j$:
\begin{align}
	P\left(j \given i \right):= \frac{\exp(\bm{u}_j \cdot \bm{v}_{i})}{Z_i}, \label{eq:prob_w2v_ng}
\end{align}
where the denominator $Z_i=\sum_{j' \in \mathcal{A}} \exp(\bm{u}_{j'} \cdot \bm{v}_{i})$ is a normalization constant, and $\mathcal{A}$ is the set of all locations.
Although {\it word2vec} generates two embedding vectors $\bm{v_i}$ and $\bm{u}_i$---which are referred to as the in-vector and out-vector, respectively, we follow convention to use the in-vector $\bm{v}_i$ as an embedding of location $i$.

Our experiments show that {\it word2vec} best explains real-world mobility flow when $w=1$ (Fig.~\ref{fig:supp:hyperparameter}), with the flow predicted by
\begin{align}
	\label{eq:flow_w2v}
	\hat{T}_{ij} = P(i) P(j\given i) = \frac{P(i)\exp(\bm{u}_j \cdot \bm{v}_{i})}{Z_i},
\end{align}
where $P(i)$ is the probability that location $i$ appears in the trajectory.
Because computing $Z_i$ is computationally demanding, multiple approximation methods have been proposed, the most popular being negative sampling~\autocite{mikolov2013word2vec}.
Negative sampling finds the embedding vectors by solving a binary classification task, in which the model classifies whether a given center-context location pair $(i, j)$ comes from the data or artificially generated ``negative'' samples.
The negative pair $(i,j)$ is sampled in proportion to $P^\gamma(j)$ with parameter $\gamma \geq 0$.

As a heuristic, the negative sampling has a bias and maximizes a different objective function~\autocite{Gutmann2010}.
When taking into account this bias, {\it word2vec} with the negative sampling fits a probability model given by
\begin{align}
	P\left(j \given i \right):= \frac{P^ \gamma(j) \exp(\bm{u}_j \cdot \bm{v}_{i})}{Z_i}, \label{eq:prob_w2v_ng0}
\end{align}
where we redefine the normalization constant as $Z_i=\sum_{j' \in \mathcal{A}} P^ \gamma (j') \exp(\bm{u}_{j'} \cdot \bm{v}_{i})$. (See Methods section for the full calculation).

Parameter $\gamma=1$ is a special choice that ensures that, when the embedding dimension is sufficiently large, there exists optimial in-vectors and out-vectors such that $\bm{u}_i = \bm{v}_i$~\autocite{levy2014neural}.
Setting $\gamma = 1$ and substituting $\bm{u}_i = \bm{v}_j$ into Eq.~\eqref{eq:prob_w2v_ng0}, the flow predicted by {\it word2vec} is given by
\begin{align}
	\hat{T}_{ij}= P(i) P(j\given i) = \frac{P(i)P(j)\exp(\bm{v}_j \cdot \bm{v}_{i})}{Z_i}. \label{eq:flow_w2v_ng1}
\end{align}
The flow $\hat{T}_{ij}$ is symmetric (\ie $\hat{T}_{ij}=\hat{T}_{ji}$) because {\it word2vec} neglects whether the context $j$ appears before or after the center $i$ in the trajectory.
If we swap $i$ and $j$ in Eq.~\eqref{eq:flow_w2v_ng1}, the numerator remains the same but the denominator can be different.
Therefore, to ensure $\hat{T}_{ij} = \hat{T}_{ji}$, the denominator $Z_i$ must be a constant.

Taken altogether, the {\it word2vec} model with the negative sampling predicts a flow in the same form as in the gravity model:
\begin{align}
	\hat{T}_{ij}=  C P(i) P(j)  \exp(\bm{v}_j \cdot \bm{v}_{i}), \label{eq:flow_w2v_ng0}
\end{align}
where $C$ is a positive constant.
In other words, \emph{word2vec with the negative sampling is a gravity model}, with the mass given by the location's frequency $P(i)$, and the distance measured by their dot similarities.
While the gravity model generates mobility flow from the given mass and locations, {\it word2vec} finds the mass and locations that best explain the given mobility flow.
Therefore, the embedding distance is inherently tied to mobility flow, and hence, has greater predictive power than the geographic distance.

A central assumption in our mathematical analysis is that the embedding vectors are optimum in terms of the {\it word2vec}'s objective function.
However, this premise is not practical because the optimization algorithm (\eg stochastic gradient descent algorithm) often finds sub-optimal embeddings.
This has two consequences.
First, the in- and out-vectors can be different.
Second, dot similarity calculated by {\it word2vec} is the distance metric tied to the gravity model, and hence, should capture the mobility flow well.
Our numerical simulations show that dot similarity calculated by {\it word2vec} is the second-best predictor of mobility flow among the ten embedding distances~(Tables~\ref{supp:table:r2_table},~\ref{supp:table:rmse_exp_table},~\ref{supp:table:rmse_power_table}).
However, cosine similarity predicted mobility flow better than dot similarity.
This result echos many applications of word embedding such as word analogy test~\autocite{levy2015improving}, in which cosine similarity outperforms dot similarity.
In predicting mobility flow, if the embedding vectors are optimal, cosine similarity should be no better than dot similarity because cosine similarity is a special case of dot similarity.
Therefore, the inferiority of dot similarity is attributed to the optimization process.
Although finding an optimal embedding is difficult in practice, a careful tuning of the optimization algorithm would close the gap between our mathematical and numerical results.

>>>>

Why does the embedding distance predict actual mobility flows better than the geographic distance?
Here, we show that the objective function of {\it word2vec} is equivalent to the gravity model, explaining its ability to learn gravity-like relationships between locations.

The {\it word2vec} model takes a location trajectory, denoted by ($a_{1}, a_{2}, \ldots, a_{T}$), as input.
A location $a_t=i$ is considered to have a context location $a_{t'}=j$ that appears in the previous or subsequent $w$ locations in the trajectory, \ie  $j \in [a_{t-w}, \ldots, a_{t-1}, a_{t+1},\ldots, a_{t+w}]$.
	{\it Word2vec} learns an embedding by estimating the probability that location $i$ has context $j$:
\begin{align}
	P\left(j \given i \right):= \frac{\exp(\bm{u}_j \cdot \bm{v}_{i})}{Z_i}, \label{eq:prob_w2v_ng}
\end{align}
where the denominator $Z_i=\sum_{j' \in \mathcal{A}} \exp(\bm{u}_{j'} \cdot \bm{v}_{i})$ is a normalization constant, and $\mathcal{A}$ is the set of all locations.
Although {\it word2vec} generates two embedding vectors $\bm{v_i}$ and $\bm{u}_i$---referred to as the in-vector and out-vector, respectively---we follow convention to use the in-vector $\bm{v}_i$ as an embedding of location $i$.

Our experiments show that {\it word2vec} best explains real-world mobility flow when $w=1$ (Fig.~\ref{fig:supp:hyperparameter}), with the flow predicted by
\begin{align}
	\label{eq:flow_w2v}
	\hat{T}_{ij} = P(i) P(j\given i) = \frac{P(i)\exp(\bm{u}_j \cdot \bm{v}_{i})}{Z_i},
\end{align}
where $P(i)$ is the probability that location $i$ appears in the trajectory.
Because computing $Z_i$ is computationally demanding, multiple approximation methods have been proposed, the most popular being negative sampling~\autocite{mikolov2013word2vec}.
Negative sampling finds the embedding vectors by solving a binary classification task, in which the model classifies whether a given target-context location pair $(i, j)$ comes from the data or from artificially generated ``negative'' samples.
The negative pair $(i,j)$ is sampled in proportion to $P^\gamma(j)$ with parameter $\gamma \geq 0$.

As a heuristic, the negative sampling has a bias and maximizes a different objective function~\autocite{Gutmann2010}.
When taking into account this bias, {\it word2vec} with the negative sampling fits a probability model given by
\begin{align}
	P\left(j \given i \right):= \frac{P^ \gamma(j) \exp(\bm{u}_j \cdot \bm{v}_{i})}{Z_i}, \label{eq:prob_w2v_ng0}
\end{align}
where we redefine the normalization constant as $Z_i=\sum_{j' \in \mathcal{A}} P^ \gamma (j') \exp(\bm{u}_{j'} \cdot \bm{v}_{i})$. (See Methods section for the full derivation).

Parameter $\gamma=1$ is a special choice that ensures that, when the embedding dimension is sufficiently large, there exists optimial in-vectors and out-vectors such that $\bm{u}_i = \bm{v}_i$~\autocite{levy2014neural}.
Setting $\gamma = 1$ and substituting $\bm{u}_i = \bm{v}_j$ into Eq.~\eqref{eq:prob_w2v_ng0}, the flow predicted by {\it word2vec} is given by
\begin{align}
	\hat{T}_{ij}= P(i) P(j\given i) = \frac{P(i)P(j)\exp(\bm{v}_j \cdot \bm{v}_{i})}{Z_i}. \label{eq:flow_w2v_ng1}
\end{align}
The flow $\hat{T}_{ij}$ is symmetric (\ie $\hat{T}_{ij}=\hat{T}_{ji}$) because {\it word2vec} neglects whether the context $j$ appears before or after the target $i$ in the trajectory.
If we swap $i$ and $j$ in Eq.~\eqref{eq:flow_w2v_ng1}, the numerator remains the same but the denominator can be different.
Therefore, to ensure $\hat{T}_{ij} = \hat{T}_{ji}$, the denominator $Z_i$ must be a constant.

Taken altogether, the {\it word2vec} model with the negative sampling predicts a flow in the same form as in the gravity model:
\begin{align}
	\hat{T}_{ij}=  C P(i) P(j)  \exp(\bm{v}_j \cdot \bm{v}_{i}), \label{eq:flow_w2v_ng0}
\end{align}
where $C$ is a positive constant.
In other words, \emph{word2vec with the negative sampling is a gravity model}, with the mass given by the location's frequency $P(i)$, and the distance measured by their dot similarities.
While the gravity model generates mobility flow from the given mass and locations, {\it word2vec} finds the mass and locations that best explain the given mobility flow.
Therefore, the embedding distance is inherently tied to the mobility flow, and hence, has greater predictive power than the geographic distance and other baselines.

A key assumption in our mathematical analysis is that the embedding vectors are optimum in terms of the {\it word2vec}'s objective function.
However, this premise is not practical because the optimization algorithm (\eg stochastic gradient descent algorithm) often finds sub-optimal embeddings.
This has two consequences.
First, the in- and out-vectors can be different.
Second, {\it word2vec} learns the dot distance between location vectors, and so this distance should best capture mobility flows. 
However, our numerical simulations show that the dot similarity between vectors is outperformed by the cosine distance as a predictor of mobility flow between scientific organizations~(Tables~\ref{supp:table:r2_table},~\ref{supp:table:rmse_exp_table},~\ref{supp:table:rmse_power_table}).
This result echos other applications of word embedding, such as word analogy testing~\autocite{levy2015improving}, in which cosine distance also outperformed dot similarity.
Given that cosine similarity is a special case of dot similarity, we speculate that the deviation in explanatory and predictive quality can be attributed to \textit{word2vec's} optimization process.
Although finding an optimal embedding is difficult in practice, careful tuning of the optimization algorithm would close the gap between our mathematical and numerical results.
